---
title: "JCEM - jc.2024-03763 - Lipid metabolism in Tangier disease"
author: "Oleksandr Petrenko, Georg Semmler"
date: "04.02.2025"
output: 
  html_document: 
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: show
  word_document: default
  pdf_document: default
---

```{r}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

# Method 1 - all
## Setup
```{r, include=FALSE}
# Load required packages
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(
  tidyverse, dplyr, readxl, ggplot2, patchwork, factoextra, gridExtra, here, 
  PCAtools, limma, EnhancedVolcano, compositions, writexl
)

# Load data
data <- read_excel("NMR_results_method_1.xlsx")

# Load Metadata
data_meta <- data.frame(Group = c("Control", "Control", "Control", "Control", 
               "Control", "Control", "Control", "Control", 
               "Control", "Control", "Disease", "Disease", 
               "Disease", "Disease"))

# Format and transform the data
data <- as.data.frame(t(data))
data <- data %>% mutate(across(where(is.character), as.numeric))

# Replace zeros to avoid log issues
data[data == 0] <- min(data[data > 0], na.rm = TRUE) * 0.5

# Log transformation and standardization
data <- log(data) +1
data <- as.data.frame(scale(data))
```

## Differential Abundance Analysis with LiMMA
```{r}
data_meta$Group <- as.factor(data_meta$Group)
design_matrix <- model.matrix(~ 0 + Group, data = data_meta)
colnames(design_matrix) <- levels(data_meta$Group)

fit <- lmFit(data, design_matrix)
contrast_matrix <- makeContrasts(Group1vsGroup2 = Disease - Control, levels = design_matrix)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

topTable_results <- topTable(fit2, adjust.method = "fdr", number = Inf)
DAA_method_1_all <- topTable_results %>%
    filter(abs(logFC) > 0.5, adj.P.Val < 0.05) %>%
    arrange(desc(logFC))

# Display filtered results & export to Excel
print(DAA_method_1_all)
write_xlsx(cbind(Metabolites = rownames(DAA_method_1_all), DAA_method_1_all), "DAA_method_1_all.xlsx")
```

### Figure 3A - Volcano plot
```{r}
# Generate Volcano Plot
volcano_plot <- EnhancedVolcano(
    DAA_method_1_all,
    lab = rownames(DAA_method_1_all),
    x = 'logFC',
    y = 'P.Value',
    xlim = c(-5, 5),
    title = 'Volcano Plot',
    legendPosition = 'top'
)
volcano_plot
```

## Prinicpal Component Analysis
```{r}
data_meta$Group <- factor(ifelse(data_meta$Group == 'Control', 'Healthy control', 'Tangier disease'), 
                          levels = c('Healthy control', 'Tangier disease'))

rownames(data_meta) <- colnames(data)

p <- pca(data, metadata = data_meta, removeVar = 0.1)

PCA_method_1_all <- as.data.frame(p$loadings)

# Display filtered results & export to Excel
print(PCA_method_1_all)
write_xlsx(cbind(Metabolites = rownames(PCA_method_1_all), PCA_method_1_all), "PCA_method_1_all.xlsx")
```

### Figure 3B - Biplot
```{r}
# Create the biplot
biplot_plot <-  
  biplot(p, 
       showLoadings = TRUE,
       lab = NULL, 
       pointSize = 3, 
       sizeLoadingsNames = 5,
       ellipse = FALSE,
       colkey = c('Tangier disease'= '#fdd635', 'Healthy control' = '#a7abed'),
       colby = 'Group', 
       legendPosition = 'top', 
       legendLabSize = 16, 
       legendIconSize = 8.0
       )
biplot_plot
```

## Safe to PDF
```{r}
pdf("Figure 3.pdf", width = 8, height = 14)  

grid.arrange(volcano_plot + ggtitle("A"), 
             biplot_plot + ggtitle("B"), 
             ncol = 1)
dev.off()
```

# Method 1 - exclude patient #4
```{r}
# Load data
data <- read_excel("NMR_results_method_1.xlsx")
data <- data[-11, ]

# Load Metadata
data_meta <- data.frame(Group = c("Control", "Control", "Control", "Control", 
               "Control", "Control", "Control", "Control", 
               "Control", "Control", "Disease", "Disease", "Disease"
               ))

# Format and transform the data
data <- as.data.frame(t(data))
data <- data %>% mutate(across(where(is.character), as.numeric))

# Replace zeros to avoid log issues
data[data == 0] <- min(data[data > 0], na.rm = TRUE) * 0.5

# Log transformation and standardization
data <- log(data) +1
data <- as.data.frame(scale(data))
```

## Differential Abundance Analysis with LiMMA
```{r}
data_meta$Group <- as.factor(data_meta$Group)
design_matrix <- model.matrix(~ 0 + Group, data = data_meta)
colnames(design_matrix) <- levels(data_meta$Group)

fit <- lmFit(data, design_matrix)
contrast_matrix <- makeContrasts(Group1vsGroup2 = Disease - Control, levels = design_matrix)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

topTable_results <- topTable(fit2, adjust.method = "fdr", number = Inf)
DAA_method_1_subgroup <- topTable_results %>%
    filter(abs(logFC) > 0.5, adj.P.Val < 0.05) %>%
    arrange(desc(logFC))

# Display filtered results & export to Excel
print(DAA_method_1_subgroup)
write_xlsx(cbind(Metabolites = rownames(DAA_method_1_subgroup), DAA_method_1_subgroup), "DAA_method_1_subgroup.xlsx")
```

### Supplementary figure 5A - Volcano plot
```{r}
# Generate Volcano Plot
volcano_plot <- EnhancedVolcano(
    DAA_method_1_subgroup,
    lab = rownames(DAA_method_1_subgroup),
    x = 'logFC',
    y = 'P.Value',
    xlim = c(-5, 5),
    title = 'Volcano Plot',
    legendPosition = 'top'
)
volcano_plot
```

## Prinicpal Component Analysis
```{r}
data_meta$Group <- factor(ifelse(data_meta$Group == 'Control', 'Healthy control', 'Tangier disease'), 
                          levels = c('Healthy control', 'Tangier disease'))

rownames(data_meta) <- colnames(data)

p <- pca(data, metadata = data_meta, removeVar = 0.1)

PCA_method_1_subgroup <- as.data.frame(p$loadings)

# Display filtered results & export to Excel
print(PCA_method_1_subgroup)
write_xlsx(cbind(Metabolites = rownames(PCA_method_1_subgroup), PCA_method_1_subgroup), "PCA_method_1_subgroup.xlsx")
```

### Supplementary figure 5B - Biplot
```{r}
# Create the biplot
biplot_plot <-  
  biplot(p, 
       showLoadings = TRUE,
       lab = NULL, 
       pointSize = 3, 
       sizeLoadingsNames = 5,
       ellipse = FALSE,
       colkey = c('Tangier disease'= '#fdd635', 'Healthy control' = '#a7abed'),
       colby = 'Group', 
       legendPosition = 'top', 
       legendLabSize = 16, 
       legendIconSize = 8.0
       )
biplot_plot
```

## Supplementary figure 5 
```{r}
tiff("Supp fig 5.tiff", width = 10, height = 12, units = "in", res = 300, compression = "lzw")

grid.arrange(volcano_plot + ggtitle("A"), 
             biplot_plot + ggtitle("B"), 
             ncol = 1)

dev.off()
```

# Method 2
```{}
# Load data
data <- read_excel("NMR_results_method_2.xlsx")

# Delete all percent or ratios
data <- data %>%
  select(-contains("Prozent"), -contains("size"), -TG_to_PG, -ApoB_to_ApoA1, -Unsaturation, -PUFA_to_MUFA, -Omega_6_to_Omega_3)

# Load Metadata
data_meta <- data.frame(Group = c("Disease", "Disease", 
               "Control", "Control", "Control", "Control", 
               "Control", "Control", "Control", "Control", 
               "Control", "Control", "Control", "Control", 
               "Control"))

# Format and transform the data
data <- as.data.frame(t(data))
data <- data %>% mutate(across(where(is.character), as.numeric))

# Replace zeros to avoid log issues
data[data == 0] <- min(data[data > 0], na.rm = TRUE) * 0.5

# Log transformation and standardization
data <- log1p(data)  # log1p(x) is log(x + 1)
data <- as.data.frame(scale(data))
```


```{r}
# Load data
data <- read_excel("NMR_results_method_2.xlsx")

# Delete all percent or ratios
data <- data %>%
  select(-contains("pct"), -contains("size"), -TG_to_PG, -ApoB_to_ApoA1, -Unsaturation, -PUFA_to_MUFA, -Omega_6_to_Omega_3)

# Load Metadata
data_meta <- data.frame(Group = c("Disease", "Disease", 
               "Control", "Control", "Control", "Control", 
               "Control", "Control", "Control", "Control", 
               "Control", "Control", "Control", "Control", 
               "Control"))

# Format and transform the data
data <- as.data.frame(t(data))
data <- data %>% mutate(across(where(is.character), as.numeric))

# Replace zeros to avoid log issues
#data[data == 0] <- min(data[data > 0], na.rm = TRUE) * 0.5

# Log transformation and standardization
data <- log(data) +1
data <- as.data.frame(scale(data))
```

## Differential Abundance Analysis with LiMMA
```{r}
data_meta$Group <- as.factor(data_meta$Group)
design_matrix <- model.matrix(~ 0 + Group, data = data_meta)
colnames(design_matrix) <- levels(data_meta$Group)

fit <- lmFit(data, design_matrix)
contrast_matrix <- makeContrasts(Group1vsGroup2 = Disease - Control, levels = design_matrix)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

topTable_results <- topTable(fit2, adjust.method = "fdr", number = Inf)
DAA_method_2 <- topTable_results %>%
    filter(abs(logFC) > 0.5, adj.P.Val < 0.05) %>%
    arrange(desc(logFC))

# Display filtered results & export to Excel
print(DAA_method_2)
write_xlsx(cbind(Metabolites = rownames(DAA_method_2), DAA_method_2), "DAA_method_2.xlsx")
```

### Supplementary figure 6A - Volcano plot
```{r}
# Generate Volcano Plot
volcano_plot <- EnhancedVolcano(
    DAA_method_2,
    lab = rownames(DAA_method_2),
    x = 'logFC',
    y = 'P.Value',
    xlim = c(-5, 5),
    title = 'Volcano Plot',
    legendPosition = 'top'
)
volcano_plot
```

## Prinicpal Component Analysis
```{r}
data_meta$Group <- factor(ifelse(data_meta$Group == 'Control', 'Healthy control', 'Tangier disease'), 
                          levels = c('Healthy control', 'Tangier disease'))

rownames(data_meta) <- colnames(data)

p <- pca(data, metadata = data_meta, removeVar = 0.1)

PCA_method_2 <- as.data.frame(p$loadings)

# Display filtered results & export to Excel
print(PCA_method_2)
write_xlsx(cbind(Metabolites = rownames(PCA_method_2), PCA_method_2), "PCA_method_2.xlsx")
```

### Supplementary figure 6B - Biplot
```{r}
# Create the biplot
biplot_plot <-  
  biplot(p, 
       showLoadings = TRUE,
       lab = NULL, 
       pointSize = 3, 
       sizeLoadingsNames = 5,
       ellipse = FALSE,
       colkey = c('Tangier disease'= '#fdd635', 'Healthy control' = '#a7abed'),
       colby = 'Group', 
       legendPosition = 'top', 
       legendLabSize = 16, 
       legendIconSize = 8.0
       )
biplot_plot
```

## Supplementary figure 6
```{r}
tiff("Supp fig 6.tiff", width = 10, height = 12, units = "in", res = 300, compression = "lzw")

grid.arrange(volcano_plot + ggtitle("A"), 
             biplot_plot + ggtitle("B"), 
             ncol = 1)

dev.off()
```
